---
title: "MIMIC-IV: AKI in Mechanically Ventilated Patients (v2 — Imputed Baseline)"
author: "Rica Pitogo"
date: "`r Sys.Date()`"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Setting up libraries
```{r include=FALSE}
library(tidyverse)
library(lubridate)
library(ggplot2)
library(scales)
library(patchwork)
library(duckdb)

mimic_path <- path.expand("~/mimic/")
```

## Loading the data
```{r}
# Core demographic and stay information

# Patients
patients <- read_csv(paste0(mimic_path, "hosp/patients.csv.gz"))

# Admissions
admissions <- read_csv(paste0(mimic_path, "hosp/admissions.csv.gz")) %>%
  filter(!is.na(admittime))

# ICU stays
icustays <- read_csv(paste0(mimic_path, "icu/icustays.csv.gz")) %>%
  filter(!is.na(intime))
```

```{r}
# ICU item dictionary
d_items <- read_csv(paste0(mimic_path, "icu/d_items.csv.gz"))

# Procedure data
procedureevents <- read_csv(paste0(mimic_path, "icu/procedureevents.csv.gz")) %>%
  filter(!is.na(starttime))

# Diagnoses data
diagnoses_icd <- read_csv(paste0(mimic_path, "hosp/diagnoses_icd.csv.gz"))
```

## Load Laboratory Data (Including Creatinine)
```{r}
# Explicit itemid → lab_category mapping (verified MIMIC-IV itemids)
# Multi-code labs:
#   lactate: 3 codes (arterial/venous/whole blood — same measurement, pooled)
#   ph/pao2/paco2: arterial blood gas only (required for P/F ratio and SOFA)
#   bilirubin: total only (51966; direct bilirubin 51464 excluded — SOFA uses total)
lab_itemids <- tribble(
  ~lab_category,   ~itemid,
  "creatinine",    50912,
  "bun",           51842,
  "potassium",     50971,
  "sodium",        50983,
  "bicarbonate",   50882,
  "hemoglobin",    51222,
  "hematocrit",    51221,
  "platelet",      51265,
  "wbc",           51301,
  "albumin",       50862,
  "lactate",       50813,
  "lactate",       52442,
  "lactate",       53154,
  "glucose",       50931,
  "chloride",      50902,
  "calcium",       50893,
  "magnesium",     50960,
  "phosphate",     50970,
  "ph",            50820,
  "pao2",          50821,
  "paco2",         50818,
  "bnp",           51921,
  "troponin",      51003,
  "bilirubin",     50885,
  "inr",           51237,
  "prothrombin",   51274,
  "ptt",           51275,
  "fibrinogen",    51214
)

# Query labevents using DuckDB
aki_itemids_str <- paste(lab_itemids$itemid, collapse = ",")
con <- dbConnect(duckdb::duckdb(), ":memory:")

aki_labs <- dbGetQuery(con, sprintf("
  SELECT
    subject_id,
    hadm_id,
    charttime,
    storetime,
    itemid,
    valuenum,
    valueuom
  FROM read_csv_auto('%shosp/labevents.csv.gz')
  WHERE itemid IN (%s)
    AND valuenum IS NOT NULL
", mimic_path, aki_itemids_str))

dbDisconnect(con, shutdown = TRUE)

# Parse and clean
aki_labs <- aki_labs %>%
  left_join(
          lab_itemids %>% select(itemid, lab_category),
          by = "itemid") %>%
  mutate(
          charttime = parse_date_time(
                  charttime, orders = c("ymd HMS", "ymd HM", "ymd"),
                  quiet = TRUE)) %>%
  filter(!is.na(charttime)) %>%
  mutate(
          storetime = parse_date_time(
                  storetime, orders = c("ymd HMS", "ymd HM", "ymd"),
                  quiet = TRUE)) %>%
  filter(!is.na(storetime))

# Extract creatinine specifically
creatinine <- aki_labs %>%
  filter(lab_category == "creatinine") %>%
  select(subject_id, hadm_id, charttime, storetime, creat = valuenum) %>%
  filter(creat > 0, creat < 20)

cat(sprintf("Loaded %s creatinine measurements\n",
            format(nrow(creatinine), big.mark=",")))

# Save for later use
saveRDS(aki_labs, "aki_labs_long.rds")
```

## Identify Mechanical Ventilation
```{r}
# Find ventilation-related procedure codes
 vent_items <- d_items %>%
  filter(
    grepl("ventilat|intubat|extubat", label, ignore.case = TRUE) |
    grepl("invasive vent", label, ignore.case = TRUE)
  ) %>%
  select(itemid, label, category)

# Get ventilation events for each ICU stay
vent_stays <- procedureevents %>%
  filter(itemid %in% vent_items$itemid) %>%
  filter(!is.na(starttime), !is.na(endtime)) %>%
  group_by(stay_id) %>%
  summarise(
    vent_start = min(starttime, na.rm = TRUE),
    vent_end = max(endtime, na.rm = TRUE),
    n_vent_events = n(),
    .groups = "drop"
  ) %>%
  mutate(
    vent_duration_hours = as.numeric(difftime(vent_end,
                                              vent_start, units = "hours"))
  ) %>%
  filter(is.finite(vent_duration_hours))

cat(sprintf("Found %s mechanically ventilated ICU stays\n",
            format(nrow(vent_stays), big.mark=",")))
```

## Build Initial Cohort (Basic Inclusion Criteria)
```{r}
cohort_step1 <- vent_stays %>%
  inner_join(icustays, by = "stay_id") %>%
  inner_join(admissions, by = c("subject_id", "hadm_id")) %>%
  inner_join(patients, by = "subject_id") %>%
  mutate(
    age = year(admittime) - anchor_year + anchor_age,
    icu_los_days = as.numeric(difftime(outtime, intime, units = "days")),
    hosp_los_days = as.numeric(difftime(dischtime, admittime, units = "days"))
  ) %>%
  filter(
    age >= 18,
    vent_duration_hours >= 24,
    !is.na(icu_los_days),
    !is.na(hosp_los_days),
    !is.na(age)
  )

cat(sprintf("Initial cohort: %s ICU stays\n",
            format(nrow(cohort_step1), big.mark=",")))
cat(sprintf("                %s unique patients\n\n",
            format(n_distinct(cohort_step1$subject_id), big.mark=",")))
```

## Exclusion #1 - ESRD/Kidney Transplant
```{r}
cat("=== EXCLUSION #1: ESRD/Kidney Transplant ===\n\n")

# ICD-9 codes (MIMIC-IV stores ICD-9 codes WITHOUT decimal points)
esrd_icd9 <- c(
  "5856",    # End stage renal disease
  "V451",    # Renal dialysis status
  "V4511",   # Renal transplant status
  "V560",    # Extracorporeal dialysis
  "V561",    # Fitting of dialysis catheter
  "V562",    # Peritoneal dialysis catheter
  "V568"     # Other dialysis
)

# ICD-10 codes
esrd_icd10 <- c(
  "N18.6",   # End stage renal disease
  "Z94.0",   # Kidney transplant status
  "Z99.2"    # Dependence on renal dialysis
)

# Find patients with these diagnoses
esrd_patients <- diagnoses_icd %>%
  filter(
    (icd_version == 9 & icd_code %in% esrd_icd9) |
    (icd_version == 10 & icd_code %in% esrd_icd10)
  ) %>%
  distinct(subject_id)

cat(sprintf("Found %s patients with ESRD/transplant\n",
            format(nrow(esrd_patients), big.mark=",")))

# Apply exclusion
cohort_step2 <- cohort_step1 %>%
  anti_join(esrd_patients, by = "subject_id")

n_excluded <- nrow(cohort_step1) - nrow(cohort_step2)

cat(sprintf("\nExcluded: %s ICU stays\n", format(n_excluded, big.mark=",")))
cat(sprintf("Remaining: %s ICU stays (%.1f%% of initial)\n\n",
            format(nrow(cohort_step2), big.mark=","),
            (nrow(cohort_step2) / nrow(cohort_step1)) * 100))
```

## Exclusion #2 - Elective Surgical Admissions
```{r}
cat("=== EXCLUSION #2: Elective Surgical Admissions ===\n\n")

cat("Admission types in current cohort:\n")
cohort_step2 %>%
  count(admission_type) %>%
  arrange(desc(n)) %>%
  mutate(percent = (n / sum(n)) * 100) %>%
  print()

cat("\nExcluding: ELECTIVE and SURGICAL SAME DAY ADMISSION\n")

cohort_step3 <- cohort_step2 %>%
  filter(!admission_type %in% c("ELECTIVE", "SURGICAL SAME DAY ADMISSION"))

n_excluded <- nrow(cohort_step2) - nrow(cohort_step3)

cat(sprintf("\nExcluded: %s ICU stays\n", format(n_excluded, big.mark=",")))
cat(sprintf("Remaining: %s ICU stays (%.1f%% of initial)\n\n",
            format(nrow(cohort_step3), big.mark=","),
            (nrow(cohort_step3) / nrow(cohort_step1)) * 100))
```

## Exclusion #3 - Death Within 48 Hours
```{r}
cat("=== EXCLUSION #3: Death Within 48 Hours of Intubation ===\n\n")

cohort_step4 <- cohort_step3 %>%
  mutate(
    hours_to_death = if_else(
      !is.na(deathtime),
      as.numeric(difftime(deathtime, vent_start, units = "hours")),
      NA_real_
    ),
    died_within_48h = !is.na(hours_to_death) &
                      hours_to_death >= 0 &
                      hours_to_death <= 48
  )

# Show breakdown
cat("Death timing:\n")
cohort_step3 %>%
  mutate(
    hours_to_death = if_else(
      !is.na(deathtime),
      as.numeric(difftime(deathtime, vent_start, units = "hours")),
      NA_real_
    ),
    died_within_48h = !is.na(hours_to_death) &
            hours_to_death >= 0 & hours_to_death <= 48
  ) %>%
  summarise(
    total_deaths = sum(!is.na(deathtime)),
    deaths_within_48h = sum(died_within_48h, na.rm = TRUE),
    deaths_after_48h = sum(!is.na(hours_to_death) &
                                   hours_to_death > 48, na.rm = TRUE)
  ) %>%
  print()

# Apply exclusion
cohort_step4 <- cohort_step4 %>%
  filter(!died_within_48h)

n_excluded <- nrow(cohort_step3) - nrow(cohort_step4)

cat(sprintf("\nExcluded: %s ICU stays\n", format(n_excluded, big.mark=",")))
cat(sprintf("Remaining: %s ICU stays (%.1f%% of initial)\n\n",
            format(nrow(cohort_step4), big.mark=","),
            (nrow(cohort_step4) / nrow(cohort_step1)) * 100))
```

## Exclusion #4 - Baseline Creatinine (Observed + Imputed) & Exclude AKI Stage 2+
```{r}
cat("=== EXCLUSION #4: Baseline Creatinine & AKI Stage 2+ ===\n\n")
cat("VERSION 2: Missing baseline creatinine is IMPUTED via CKD-EPI 2021\n")
cat("           back-solved at assumed GFR = 90 mL/min/1.73m²\n\n")

#============================================================================
# STEP 4A: Calculate Observed Baseline Creatinine
# Definition: Median creatinine from 7-365 days before admission
#============================================================================

cat("Step 4A: Calculating baseline creatinine (observed + imputed)...\n")
cat("Observed: MEDIAN creatinine from 7-365 days before hospital admission\n")
cat("Imputed:  CKD-EPI 2021 back-calculation at GFR = 90 (age + sex)\n\n")

# --- 4A-i: Observed baseline ---
observed_baseline <- cohort_step4 %>%
  select(subject_id, hadm_id, stay_id, admittime) %>%
  inner_join(
    creatinine %>% select(subject_id, charttime, creat),
    by = "subject_id",
    relationship = "many-to-many"
  ) %>%
  mutate(
    days_before_admit = as.numeric(difftime(admittime, charttime,
                                            units = "days"))
  ) %>%
  filter(
    days_before_admit >= 7,
    days_before_admit <= 365
  ) %>%
  group_by(stay_id) %>%
  summarise(
    baseline_creat       = median(creat, na.rm = TRUE),
    n_baseline_measures  = n(),
    earliest_measure_days = max(days_before_admit, na.rm = TRUE),
    latest_measure_days   = min(days_before_admit, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(baseline_source = "observed")

cat(sprintf("  Stays with OBSERVED baseline creatinine: %s / %s (%.1f%%)\n",
            format(nrow(observed_baseline), big.mark = ","),
            format(nrow(cohort_step4),      big.mark = ","),
            nrow(observed_baseline) / nrow(cohort_step4) * 100))

# --- 4A-ii: CKD-EPI 2021 back-calculation function ---
# GFR = 142 * min(Scr/k,1)^alpha * max(Scr/k,1)^-1.2 * 0.9938^Age * (1.012 if F)
# Back-solve for Scr at GFR = 90
impute_baseline_creat_ckdepi <- function(age, gender, gfr = 90) {
  female    <- gender == "F"
  kappa     <- ifelse(female, 0.7,    0.9)
  alpha     <- ifelse(female, -0.241, -0.302)
  f_fac     <- ifelse(female, 1.012,  1.0)

  gfr_adj   <- gfr / (142 * 0.9938^age * f_fac)

  # Case 1: Scr <= kappa  →  Scr = kappa * GFR_adj^(1/alpha)
  scr_case1 <- kappa * gfr_adj^(1 / alpha)
  # Case 2: Scr > kappa   →  Scr = kappa * GFR_adj^(-1/1.2)
  scr_case2 <- kappa * gfr_adj^(-1 / 1.2)

  # Pick self-consistent solution
  ifelse(scr_case1 <= kappa, scr_case1, scr_case2)
}

# --- 4A-iii: Impute for stays missing observed baseline ---
stays_missing_baseline <- cohort_step4 %>%
  anti_join(observed_baseline, by = "stay_id") %>%
  select(stay_id, age, gender)

imputed_baseline <- stays_missing_baseline %>%
  mutate(
    baseline_creat      = impute_baseline_creat_ckdepi(age, gender, gfr = 90),
    n_baseline_measures = 0L,
    baseline_source     = "imputed"
  ) %>%
  select(stay_id, baseline_creat, n_baseline_measures, baseline_source)

cat(sprintf("  Stays with IMPUTED baseline creatinine: %s (%.1f%%)\n\n",
            format(nrow(imputed_baseline), big.mark = ","),
            nrow(imputed_baseline) / nrow(cohort_step4) * 100))

# --- 4A-iv: Combine observed + imputed ---
baseline_creatinine <- bind_rows(
  observed_baseline %>% select(stay_id, baseline_creat,
                                n_baseline_measures, baseline_source),
  imputed_baseline
)

cat("Baseline creatinine summary by source:\n")
baseline_creatinine %>%
  group_by(baseline_source) %>%
  summarise(
    n       = n(),
    mean_cr = round(mean(baseline_creat, na.rm = TRUE), 3),
    median_cr = round(median(baseline_creat, na.rm = TRUE), 3),
    sd_cr   = round(sd(baseline_creat, na.rm = TRUE), 3),
    .groups = "drop"
  ) %>%
  print()

#============================================================================
# STEP 4B: All stays now have baseline — report but do not exclude
#============================================================================

cat("\nStep 4B: Baseline creatinine coverage check...\n")
cat("  (In v2, missing observed baseline is imputed — no stays excluded here)\n")

n_still_missing <- nrow(cohort_step4) - nrow(baseline_creatinine)
cat(sprintf("  Stays still missing baseline (e.g., missing age/gender): %s\n",
            format(n_still_missing, big.mark = ",")))

cohort_step4_with_baseline <- cohort_step4 %>%
  semi_join(baseline_creatinine, by = "stay_id")

cat(sprintf("  Remaining: %s stays\n\n",
            format(nrow(cohort_step4_with_baseline), big.mark = ",")))

#============================================================================
# STEP 4C: Get Peri-Intubation Creatinine
#============================================================================

cat("Step 4C: Finding creatinine at intubation...\n")
cat("  Window: 48 hours before to 6 hours after intubation\n\n")

peri_intubation_creat <- cohort_step4_with_baseline %>%
  select(subject_id, hadm_id, stay_id, vent_start) %>%
  inner_join(
    creatinine,
    by = c("subject_id", "hadm_id"),
    relationship = "many-to-many"
  ) %>%
  mutate(
    hours_from_vent = as.numeric(difftime(charttime, vent_start,
                                          units = "hours"))
  ) %>%
  filter(
    hours_from_vent >= -48,
    hours_from_vent <= 6
  ) %>%
  group_by(stay_id) %>%
  summarise(
    max_creat_at_intubation = max(creat, na.rm = TRUE),
    min_creat_at_intubation = min(creat, na.rm = TRUE),
    n_measures = n(),
    .groups = "drop"
  )

cat(sprintf("  Stays with peri-intubation creatinine: %s / %s (%.1f%%)\n\n",
            format(nrow(peri_intubation_creat), big.mark = ","),
            format(nrow(cohort_step4_with_baseline), big.mark = ","),
            nrow(peri_intubation_creat) / nrow(cohort_step4_with_baseline) * 100))

#============================================================================
# STEP 4C.5: Require Peri-Intubation Creatinine
#============================================================================

cat("Step 4C.5: Requiring peri-intubation creatinine data...\n")

cohort_step4_with_peri <- cohort_step4_with_baseline %>%
  semi_join(peri_intubation_creat, by = "stay_id")

n_excluded_no_peri <- nrow(cohort_step4_with_baseline) -
        nrow(cohort_step4_with_peri)
cat(sprintf("  Excluded (no peri-intubation creatinine): %s stays\n",
            format(n_excluded_no_peri, big.mark = ",")))
cat(sprintf("  Remaining: %s stays\n\n",
            format(nrow(cohort_step4_with_peri), big.mark = ",")))

#============================================================================
# STEP 4D: Stage AKI Using KDIGO Criteria
#============================================================================

cat("Step 4D: Staging AKI at intubation (KDIGO criteria)...\n\n")

aki_at_intubation <- baseline_creatinine %>%
  inner_join(peri_intubation_creat, by = "stay_id") %>%
  mutate(
    creat_fold_increase = max_creat_at_intubation / baseline_creat,

    aki_stage_creat = case_when(
      creat_fold_increase >= 3.0 ~ "Stage 3",
      creat_fold_increase >= 2.0 ~ "Stage 2",
      creat_fold_increase >= 1.5 ~ "Stage 1",
      TRUE ~ "No AKI"
    ),

    has_aki_stage2_plus = creat_fold_increase >= 2.0,
    aki_stage = case_when(
      creat_fold_increase >= 3.0 ~ "Stage 3",
      has_aki_stage2_plus           ~ "Stage 2",
      creat_fold_increase >= 1.5    ~ "Stage 1",
      TRUE                          ~ "No AKI"
    )
  )

cat("  AKI staging distribution (all stays with peri-intubation creatinine):\n")
aki_summary <- aki_at_intubation %>%
  count(aki_stage) %>%
  mutate(
    percent = (n / sum(n)) * 100,
    aki_stage = factor(aki_stage,
                       levels = c("No AKI", "Stage 1", "Stage 2", "Stage 3"))
  ) %>%
  arrange(aki_stage)

print(aki_summary)

cat("\n  AKI staging by baseline source:\n")
aki_at_intubation %>%
  count(baseline_source, aki_stage) %>%
  mutate(aki_stage = factor(aki_stage,
                            levels = c("No AKI", "Stage 1", "Stage 2", "Stage 3"))) %>%
  arrange(baseline_source, aki_stage) %>%
  print()

#============================================================================
# STEP 4E: Exclude AKI Stage 2+
#============================================================================

cat("\nStep 4E: Excluding AKI Stage 2+...\n")

stays_with_aki_stage2_plus <- aki_at_intubation %>%
  filter(has_aki_stage2_plus) %>%
  select(stay_id)

cat(sprintf("  Identified %s stays with AKI Stage 2+\n",
            format(nrow(stays_with_aki_stage2_plus), big.mark = ",")))

cohort_final <- cohort_step4_with_peri %>%
  anti_join(stays_with_aki_stage2_plus, by = "stay_id")

n_excluded_aki <- nrow(cohort_step4_with_peri) - nrow(cohort_final)

cat(sprintf("\n  Excluded: %s ICU stays\n", format(n_excluded_aki, big.mark = ",")))
cat(sprintf("  FINAL COHORT: %s ICU stays (%.1f%% of initial)\n\n",
            format(nrow(cohort_final), big.mark = ","),
            (nrow(cohort_final) / nrow(cohort_step1)) * 100))

#============================================================================
# Verification
#============================================================================

cat("=== Verification ===\n")

final_with_aki_stage <- cohort_final %>%
  left_join(aki_at_intubation %>%
                    select(stay_id, aki_stage, baseline_creat,
                           baseline_source, max_creat_at_intubation,
                           creat_fold_increase, n_baseline_measures),
            by = "stay_id")

cat("AKI stage distribution in FINAL cohort:\n")
final_with_aki_stage %>%
  count(aki_stage) %>%
  mutate(percent = (n / sum(n)) * 100) %>%
  print()

cat("\nBaseline source in FINAL cohort:\n")
final_with_aki_stage %>%
  count(baseline_source) %>%
  mutate(percent = (n / sum(n)) * 100) %>%
  print()

cat(sprintf("\nTotal in final cohort: %s\n", format(nrow(cohort_final), big.mark = ",")))
cat(sprintf("All have baseline creatinine: %s\n",
            all(!is.na(final_with_aki_stage$baseline_creat))))
cat(sprintf("All have No AKI or Stage 1 only: %s\n\n",
            all(final_with_aki_stage$aki_stage %in% c("No AKI", "Stage 1"))))
```

## Final Cohort Summary Statistics
```{r}
# Rename for consistency
cohort <- cohort_final

# Calculate summary statistics
cohort_summary <- cohort %>%
  summarise(
    N_ICU_Stays = n(),
    N_Unique_Patients = n_distinct(subject_id),
    Mean_Age = mean(age, na.rm = TRUE),
    SD_Age = sd(age, na.rm = TRUE),
    Median_Age = median(age, na.rm = TRUE),
    Percent_Female = mean(gender == "F", na.rm = TRUE) * 100,
    Median_ICU_LOS_Days = median(icu_los_days, na.rm = TRUE),
    IQR_ICU_LOS_Days = paste0(
      round(quantile(icu_los_days, 0.25, na.rm = TRUE), 1),
      " - ",
      round(quantile(icu_los_days, 0.75, na.rm = TRUE), 1)
    ),
    Median_Vent_Hours = median(vent_duration_hours, na.rm = TRUE),
    IQR_Vent_Hours = paste0(
      round(quantile(vent_duration_hours, 0.25, na.rm = TRUE), 1),
      " - ",
      round(quantile(vent_duration_hours, 0.75, na.rm = TRUE), 1)
    ),
    Hospital_Mortality_Percent = mean(hospital_expire_flag == 1,
                                      na.rm = TRUE) * 100
  )

# Print nicely
cat("Sample Size:\n")
cat(sprintf("  ICU Stays: %s\n", format(cohort_summary$N_ICU_Stays, big.mark = ",")))
cat(sprintf("  Unique Patients: %s\n\n",
            format(cohort_summary$N_Unique_Patients, big.mark = ",")))

cat("Demographics:\n")
cat(sprintf("  Age: %.1f ± %.1f years (median: %.1f)\n",
            cohort_summary$Mean_Age, cohort_summary$SD_Age,
            cohort_summary$Median_Age))
cat(sprintf("  Female: %.1f%%\n\n", cohort_summary$Percent_Female))

cat("Clinical Characteristics:\n")
cat(sprintf("  ICU Length of Stay: %.1f days (IQR: %s)\n",
            cohort_summary$Median_ICU_LOS_Days,
            cohort_summary$IQR_ICU_LOS_Days))
cat(sprintf("  Ventilation Duration: %.1f hours (IQR: %s)\n",
            cohort_summary$Median_Vent_Hours, cohort_summary$IQR_Vent_Hours))
cat(sprintf("  Hospital Mortality: %.1f%%\n\n",
            cohort_summary$Hospital_Mortality_Percent))

# Save cohort
saveRDS(cohort, "final_cohort_imputed_baseline.rds")
cat("Final cohort saved to 'final_cohort_imputed_baseline.rds'\n")
```

## Visualizations

### 1. Exclusion Flow Diagram
```{r exclusion-flow, fig.width=10, fig.height=7}
# Create exclusion flow data
exclusion_flow <- data.frame(
  Step = c(
    "Initial cohort\n(≥18y, ≥24h vent)",
    "After excluding\nESRD/transplant",
    "After excluding\nelective surgical",
    "After excluding\ndeath <48h",
    "After baseline Cr\n(observed + imputed)",
    "After requiring\nperi-intubation creatinine",
    "FINAL COHORT\n(after AKI Stage 2+ exclusion)"
  ),
  N_Stays = c(
    nrow(cohort_step1),
    nrow(cohort_step2),
    nrow(cohort_step3),
    nrow(cohort_step4),
    nrow(cohort_step4_with_baseline),
    nrow(cohort_step4_with_peri),
    nrow(cohort_final)
  )
) %>%
  mutate(
    N_Excluded = c(
      0,
      nrow(cohort_step1) - nrow(cohort_step2),
      nrow(cohort_step2) - nrow(cohort_step3),
      nrow(cohort_step3) - nrow(cohort_step4),
      nrow(cohort_step4) - nrow(cohort_step4_with_baseline),
      nrow(cohort_step4_with_baseline) - nrow(cohort_step4_with_peri),
      nrow(cohort_step4_with_peri) - nrow(cohort_final)
    ),
    Percent_Remaining = (N_Stays / nrow(cohort_step1)) * 100,
    Step = factor(Step, levels = Step)
  )

# Create waterfall chart
p_flow <- ggplot(exclusion_flow, aes(x = Step, y = N_Stays)) +
  geom_bar(stat = "identity",
           fill = c("#3498db", "#95a5a6", "#95a5a6", "#95a5a6",
                    "#95a5a6", "#95a5a6", "#2ecc71"),
           color = NA,
           width = 0.7) +
  geom_text(aes(label = format(N_Stays, big.mark = ",")),
            vjust = -0.5, size = 4.5, fontface = "bold") +
  geom_text(aes(label = sprintf("Excluded:\n%s", format(N_Excluded, big.mark = ","))),
            vjust = 1.5, size = 3, color = "red",
            data = exclusion_flow %>% filter(N_Excluded > 0)) +
  geom_text(aes(label = sprintf("(%.1f%%)", Percent_Remaining)),
            vjust = -2.5, size = 3.5, color = "darkblue") +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.18))) +
  labs(
    title = "Cohort Selection Flow — v2: Imputed Baseline Creatinine",
    subtitle = sprintf("From %s to %s ICU stays (%.1f%% retained) | Missing baseline Cr imputed via CKD-EPI 2021 at GFR=90",
                      format(nrow(cohort_step1), big.mark = ","),
                      format(nrow(cohort_final), big.mark = ","),
                      (nrow(cohort_final) / nrow(cohort_step1)) * 100),
    x = NULL,
    y = "Number of ICU Stays"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
    plot.subtitle = element_text(size = 10, hjust = 0.5),
    axis.text.x = element_text(angle = 35, hjust = 1, vjust = 1, size = 9),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.margin = margin(t = 5, r = 10, b = 20, l = 10)
  )

print(p_flow)
ggsave("exclusion_flow_diagram_v2_imputed.png", p_flow, width = 10, height = 7, dpi = 300)
```

### 2. Imputed vs Observed Baseline Creatinine Distribution
```{r baseline-source-viz, fig.width=10, fig.height=5}
final_with_aki_stage %>%
  ggplot(aes(x = baseline_creat, fill = baseline_source)) +
  geom_histogram(bins = 50, alpha = 0.7, color = "white", linewidth = 0.2,
                 position = "identity") +
  scale_fill_manual(values = c("observed" = "#3498db", "imputed" = "#e67e22"),
                    labels = c("observed" = "Observed (measured lab)",
                               "imputed"  = "Imputed (CKD-EPI at GFR=90)")) +
  scale_x_continuous(limits = c(0, 3), breaks = seq(0, 3, 0.5)) +
  labs(
    title = "Baseline Creatinine Distribution by Source",
    subtitle = sprintf("Final cohort: %s stays (%s observed, %s imputed)",
                      format(nrow(final_with_aki_stage), big.mark = ","),
                      format(sum(final_with_aki_stage$baseline_source == "observed",
                                 na.rm = TRUE), big.mark = ","),
                      format(sum(final_with_aki_stage$baseline_source == "imputed",
                                 na.rm = TRUE), big.mark = ",")),
    x = "Baseline Creatinine (mg/dL)",
    y = "Count",
    fill = "Source"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "bottom"
  )
```

## Saving Intermediate Datasets
```{r}
# Save baseline creatinine data (includes baseline_source column)
saveRDS(baseline_creatinine, "baseline_creatinine_imputed.rds")
cat("baseline_creatinine_imputed.rds\n")

# Save AKI staging data
saveRDS(aki_at_intubation, "aki_staging_at_intubation_imputed.rds")
cat("aki_staging_at_intubation_imputed.rds\n")

# Save exclusion lists
exclusion_lists_v2 <- list(
  esrd_patients = esrd_patients,
  stays_with_aki_stage2_plus = stays_with_aki_stage2_plus
)
saveRDS(exclusion_lists_v2, "exclusion_patient_lists_v2_imputed.rds")
cat("exclusion_patient_lists_v2_imputed.rds\n")

cat(sprintf("\nFinal cohort: %s ICU stays from %s patients\n",
            format(nrow(cohort), big.mark = ","),
            format(n_distinct(cohort$subject_id), big.mark = ",")))
cat("Baseline method: Observed median Cr 7-365d OR CKD-EPI 2021 imputed at GFR=90\n")
cat(sprintf("  Observed: %s | Imputed: %s\n",
            format(sum(baseline_creatinine$baseline_source == "observed"), big.mark=","),
            format(sum(baseline_creatinine$baseline_source == "imputed"), big.mark=",")))
```
