---
title: "AKI Feature Engineering — MICE Imputation"
author: "Rica Pitogo"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 9, fig.height = 5)
```

## Overview

This notebook performs MICE (Multiple Imputation by Chained Equations) on the
pre-imputation feature matrix produced by `02_feature_engineering_v3_mice.Rmd`.

**Input:** `feature_matrix_preimputation.rds` — NAs intact, correlated and zero-variance features dropped

**Outputs:**
- `feature_matrix_imputed_stacked.rds` — all 5 imputed datasets stacked (`.imp` = 1–5); use to train 5 models and average predictions
- `feature_matrix_mids.rds` — raw `mids` object for pooling-based inference

**Imputation strategy:**
- `pmm` (predictive mean matching) for all continuous columns with ≤50% missingness
- `logreg` for binary: `gender`
- `polyreg` for multi-level factors: `race`, `admission_type`
- Columns with **>50% missingness** are dropped entirely — too sparse to be useful predictors
- IDs, timestamps, outcomes, structural zeros, and SOFA-derived columns are excluded from imputation

---

## Libraries

```{r libraries}
library(tidyverse)
library(mice)
library(knitr)
```

---

## Step 1: Load Pre-Imputation Matrix

```{r load}
fm <- readRDS("feature_matrix_preimputation.rds")

cat(sprintf("Loaded: %d rows x %d cols\n", nrow(fm), ncol(fm)))
```

```{r missingness-overview}
miss_pre <- fm |>
  summarise(across(everything(), \(x) mean(is.na(x)))) |>
  pivot_longer(everything(), names_to = "variable", 
               values_to = "pct_missing") |>
  mutate(pct_missing = round(pct_missing * 100, 1)) |>
  filter(pct_missing > 0) |>
  arrange(desc(pct_missing))

cat(sprintf("%d of %d columns have missing values\n",
            nrow(miss_pre), ncol(fm)))

miss_pre |>
  head(20) |>
  kable(col.names = c("Variable", "% Missing"),
        caption = "Top 20 columns by missingness (pre-imputation)")
```

---

## Step 2: Define Imputation Roles

Columns excluded from MICE fall into four categories:
1. **IDs & timestamps** — not predictors
2. **Outcomes** — must not be used in imputation model
3. **Structural zeros** — 0 = "not given/not present" (vasopressors, comorbidities, vaso doses)
4. **Derived SOFA components** — computed from inputs; impute inputs, not the derived scores
5. **Count variables (`_n`)** — number of measurements; NA means 0 measurements taken, not missing data


```{r define-roles}
exclude_cols <- c(
  # IDs and timestamps
  "stay_id", "subject_id", "hadm_id", "anchor_year",
  "vent_start", "vent_end", "intime", "admittime", "outtime", "deathtime",
  
  # Outcomes
  "aki_stage2_7d", "time_to_aki_hours", "n_creat_obs_7d",
  
  # Structural zeros / flags
  "vasopressor_24h",
  "total_fluid_in_24h", "total_fluid_out_24h", "fluid_balance_24h",
  "dm", "htn", "ckd", "copd", "hf", "cad", "sepsis",
  "hospital_expire_flag",
  
  # Vasopressor dose rates (0 = not given)
  grep("^vaso_dose_", names(fm), value = TRUE),
  
  # SOFA components (derived; impute inputs instead)
  grep("^sofa_", names(fm), value = TRUE),

  # Count variables (already zero-filled above; 0 = not measured)
  grep("_n$", names(fm), value = TRUE)
)

# Only keep exclude_cols that actually exist in fm
exclude_cols <- intersect(exclude_cols, names(fm))

impute_cols <- setdiff(names(fm), exclude_cols)

cat(sprintf("Columns to impute: %d\n", length(impute_cols)))
cat(sprintf("Columns excluded:  %d\n", length(exclude_cols)))
```

```{r impute-cols-table}
# How many impute_cols have any missingness?
miss_impute <- fm |>
  select(all_of(impute_cols)) |>
  summarise(across(everything(), \(x) mean(is.na(x)))) |>
  pivot_longer(everything(), names_to = "variable", 
               values_to = "pct_missing") |>
  mutate(pct_missing = round(pct_missing * 100, 1)) |>
  filter(pct_missing > 0) |>
  arrange(desc(pct_missing))

cat(sprintf("%d of %d imputable columns have missing values\n",
            nrow(miss_impute), length(impute_cols)))
```

---

## Step 3: Drop Sparse and Near-Zero-Variance Columns, Build Predictor Matrix

First drop columns with >50% missingness (too sparse to be useful predictors),
then drop near-zero-variance columns identified in `03_zero_variance.Rmd`
exploration (low clinical signal in this mechanically ventilated cohort),
then build the predictor matrix on the remaining columns.

```{r quickpred}
fm_for_mice <- fm[, impute_cols]

# Convert character columns to factors for MICE
fm_for_mice <- fm_for_mice |>
  mutate(across(where(is.character), as.factor))

# Drop columns with >50% missingness BEFORE quickpred
pct_miss <- colMeans(is.na(fm_for_mice)) * 100
drop_cols <- names(pct_miss[pct_miss > 50])

cat(sprintf("%d columns dropped (>50%% missing):\n", length(drop_cols)))
if (length(drop_cols) > 0) {
  print(data.frame(variable    = drop_cols,
                   pct_missing = round(pct_miss[drop_cols], 1)),
        row.names = FALSE)
  fm_for_mice <- fm_for_mice |> select(-all_of(drop_cols))
} else {
  cat("None.\n")
}
cat(sprintf("\nRetained for imputation: %d columns\n", ncol(fm_for_mice)))

# Drop near-zero-variance columns identified in 03_zero_variance.Rmd
# gcs_verbal_* nearly constant (=1T, intubated); tidal_vol_set_sd mostly 0
nzv_drop <- c("gcs_verbal_sd", "gcs_verbal_min", "gcs_verbal_mean",
              "tidal_vol_set_sd")
nzv_drop <- intersect(nzv_drop, names(fm_for_mice))

cat(sprintf("\n%d near-zero-variance columns dropped: %s\n",
            length(nzv_drop), paste(nzv_drop, collapse = ", ")))

fm_for_mice <- fm_for_mice |> select(-all_of(nzv_drop))
cat(sprintf("Retained after NZV filter: %d columns\n", ncol(fm_for_mice)))

# Build predictor matrix on the filtered dataset
pred_mat <- quickpred(fm_for_mice, mincor = 0.1, minpuc = 0.25)

cat(sprintf("Predictor matrix: %d x %d\n", nrow(pred_mat), ncol(pred_mat)))
cat(sprintf("Avg predictors per target: %.1f\n", mean(rowSums(pred_mat))))
```

---

## Step 4: Run MICE

First pass at `maxit = 0` to get the `method` vector, then set methods before
the full run.

```{r mice-run, cache=TRUE}
# First pass — just to get the method vector
mids_init <- mice(fm_for_mice, m = 1, maxit = 0,
                  predictorMatrix = pred_mat,
                  printFlag = FALSE)

# Set methods
meth <- mids_init$method
meth[names(meth) != ""] <- "pmm" 
meth["gender"] <- "logreg"
meth["race"] <- "polyreg"
meth["admission_type"] <- "polyreg"
meth[colSums(is.na(fm_for_mice)) == 0] <- ""     

cat("Methods assigned:\n")
print(meth[meth != ""])

# Final MICE run: m=5 imputations, 10 iterations
set.seed(203)
mids_obj <- mice(
  fm_for_mice,
  m = 5,
  method = meth,
  predictorMatrix = pred_mat,
  maxit = 10,
  printFlag = FALSE
)

cat("\nMICE complete.\n")
if (!is.null(mids_obj$loggedEvents)) {
  cat(sprintf("Logged events: %d \n",
              nrow(mids_obj$loggedEvents)))
} else {
  cat("No logged events \n")
}

# Define here so it's available for Steps 5, 6, and 8
imputed_cols <- names(meth[meth != ""])
cat(sprintf("Columns imputed: %d\n", length(imputed_cols)))
```

---

## Step 5: Convergence Diagnostics

Trace plots show the mean and SD of imputed values across iterations (chains).
Good mixing = lines that wander without systematic drift.

```{r convergence-plot, fig.height=10, fig.width=12}
# Select top 8 most-missing variables that were imputed
top_miss_vars <- miss_impute |>
  filter(variable %in% imputed_cols) |>
  slice_max(pct_missing, n = 8) |>
  pull(variable)

plot(mids_obj, y = top_miss_vars, layout = c(2, length(top_miss_vars)))
```

---

## Step 6: Validate Distributions

Density overlay: observed data (blue) vs. each of the 5 imputed datasets (red).
Imputed values should follow a plausible distribution relative to observed.

```{r density-plot, fig.height=8, fig.width=10}
density_vars <- miss_impute |>
  filter(variable %in% imputed_cols, pct_missing >= 10) |>
  slice_max(pct_missing, n = 6) |>
  pull(variable)

if (length(density_vars) > 0) {
  fml <- as.formula(paste("~", paste(density_vars, collapse = " + ")))
  densityplot(mids_obj, fml,
              main = "Observed (blue) vs. Imputed (red) distributions")
} else {
  cat("No variables with ≥10% missingness — skipping density plot.\n")
}
```

---

## Step 7: Stack All 5 Imputed Datasets

Re-attach the excluded columns (IDs, outcomes, structural zeros) and stack all 5
imputations. The `.imp` column (1–5) identifies which imputation each row belongs to.

```{r stack-imputed}
meta <- fm[, c("stay_id", setdiff(exclude_cols, "stay_id")), drop = FALSE]

feature_matrix_imputed_stacked <- map_dfr(1:5, function(i) {
  complete(mids_obj, action = i) |>
    bind_cols(meta) |>
    mutate(
      .imp = i,
      # Recompute sofa_resp from imputed pf_ratio (overwrites pre-MICE NA)
      sofa_resp = case_when(
        pf_ratio < 100 ~ 4L, pf_ratio < 200 ~ 3L,
        pf_ratio < 300 ~ 2L, pf_ratio < 400 ~ 1L,
        !is.na(pf_ratio) ~ 0L, TRUE ~ NA_integer_
      )
    )
})

# Sanity checks
stopifnot(nrow(feature_matrix_imputed_stacked) == nrow(fm) * 5)
stopifnot(all(feature_matrix_imputed_stacked$.imp %in% 1:5))

cat(sprintf("Stacked dataset: %s rows x %d cols (.imp 1–5)\n",
            format(nrow(feature_matrix_imputed_stacked), big.mark = ","),
            ncol(feature_matrix_imputed_stacked)))
```

---

## Step 8: Missingness Confirmation

```{r missingness-check}
# These should be fully filled (imputed_cols defined in Step 4)
miss_after <- feature_matrix_imputed_stacked |>
  filter(.imp == 1) |>
  summarise(across(all_of(imputed_cols), \(x) sum(is.na(x)))) |>
  pivot_longer(everything(), names_to = "variable", values_to = "n_missing") |>
  filter(n_missing > 0)

if (nrow(miss_after) == 0) {
  cat("All imputed columns are fully filled — no remaining NAs.\n")
} else {
  cat(sprintf("WARNING: %d imputed columns still have NAs:\n", 
              nrow(miss_after)))
  print(miss_after, row.names = FALSE)
}

cat(sprintf("\n%d columns were dropped before imputation (>50%% missing).\n",
            length(drop_cols)))

# Full missingness summary on .imp == 1
miss_final <- feature_matrix_imputed_stacked |>
  filter(.imp == 1) |>
  summarise(across(everything(), \(x) mean(is.na(x)))) |>
  pivot_longer(everything(), 
               names_to = "variable", values_to = "pct_missing") |>
  mutate(pct_missing = round(pct_missing * 100, 1)) |>
  filter(pct_missing > 0) |>
  arrange(desc(pct_missing))

if (nrow(miss_final) > 0) {
  cat(sprintf("\nRemaining NAs post-imputation (%d columns):\n", 
              nrow(miss_final)))
  print(miss_final, n = Inf, row.names = FALSE)
}
```

---

## Step 9: Save Outputs

```{r save}
saveRDS(feature_matrix_imputed_stacked, "feature_matrix_imputed_stacked.rds")
saveRDS(mids_obj, "feature_matrix_mids.rds")

cat(sprintf("Saved feature_matrix_imputed_stacked.rds — %s rows x %d cols\n",
            format(nrow(feature_matrix_imputed_stacked), big.mark = ","),
            ncol(feature_matrix_imputed_stacked)))
cat(sprintf("Saved feature_matrix_mids.rds — mids object (m=5, maxit=10)\n"))
```
