---
title: "AKI Prediction — Machine Learning Pipeline"
subtitle: "BIOSTAT 203B Final Project"
author: "Rica Pitogo"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
    fig_width: 10
    fig_height: 6
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  cache   = FALSE,
  fig.width  = 10,
  fig.height = 6
)
```

## Overview

This notebook trains five machine learning classifiers to predict **AKI Stage 2+ within 7 days of mechanical ventilation** in ICU patients. The input is the MICE-imputed stacked feature matrix (`feature_matrix_imputed_stacked.rds`, m = 5 imputations).

**Multi-imputation (MI) strategy:**

1. Split unique `stay_id`s (80/20, stratified by outcome) — done once before touching stacked data to prevent patient-level leakage.
2. Tune hyperparameters on imputation 1 (`.imp == 1`) via 5-fold CV.
3. Fit each model on all 5 imputed training sets (20 total fits).
4. For each test patient, average predicted probabilities across 5 imputed test rows (Rubin's rule adapted for predictions).
5. Evaluate pooled predictions.

---

## 1. Libraries and Setup

```{r libraries}
library(tidyverse)
library(tidymodels)
library(ranger)      
library(xgboost)     
library(glmnet)      
library(kernlab)    
library(probably)    
library(pROC)        
library(patchwork)   
library(doParallel)  
library(knitr)
library(scales)

tidymodels_prefer()

# Set random seed
set.seed(203)
```

---

## 2. Load Data and Define Predictors

```{r load-data}
stacked <- readRDS("feature_matrix_imputed_stacked.rds")

cat("Dimensions:", nrow(stacked), "rows x", ncol(stacked), "cols\n")
cat("Imputations:", sort(unique(stacked$.imp)), "\n")
cat("Unique stay_ids:", n_distinct(stacked$stay_id), "\n")
```

```{r define-predictors}
# All 74 candidate predictors

predictor_cols <- c(
  # Demographics
  "age", "gender", "admission_type",
 
  # Labs (24h mean)
  "bicarbonate_mean", "chloride_mean", "glucose_mean", 
  "hemoglobin_mean", "inr_mean", "magnesium_mean", "paco2_mean",
  "pao2_mean", "ph_mean", "platelet_mean", "potassium_mean",
  "ptt_mean", "sodium_mean", "wbc_mean", "bilirubin_mean", "calcium_mean",
  "lactate_mean", "phosphate_mean",
  
  # Ventilator / vitals (24h mean)
  "fio2_mean", "gcs_eye_mean", "gcs_motor_mean",
  "heart_rate_mean", "map_noninvasive_mean",
  "plateau_pressure_mean", "resp_rate_obs_mean", "resp_rate_set_mean",
  "sbp_invasive_mean", "sbp_noninvasive_mean",
  "spo2_mean", "temp_f_mean", "tidal_vol_obs_mean", "tidal_vol_set_mean",
  
  # Derived ventilator
  "peep_mean", "driving_pressure_mean", "map_mean", "gcs_total_mean",
  
  # Fluids
  "total_fluid_in_24h", "total_fluid_out_24h", "fluid_balance_24h",
  
  # Vasopressors
  "vasopressor_24h", "vasopressor_hours_24h", "vaso_dose_221906", 
  "vaso_dose_221289", "vaso_dose_221653", "vaso_dose_221662",
  
  # Height and Weight
  "height_cm", "weight_kg",
  
  # Comorbidities
  "dm", "htn", "ckd", "copd", "hf", "cad", "sepsis",
  
  # Demographics
  "race",
  
  # Derived
  "bmi", "pf_ratio",
  
  # SOFA components (sofa_renal excluded)
  "sofa_resp", "sofa_coag", "sofa_liver", "sofa_cardio", "sofa_neuro"
)

predictor_cols <- intersect(predictor_cols, names(stacked))
```

---

## 3. Prepare Data

```{r prepare-data}
race_levels <- c("White", "Black", "Hispanic", "Asian", "Other/Unknown")

stacked_clean <- stacked |>
  filter(!is.na(aki_stage2_7d)) |>
  mutate(
    # Outcome: AKI as first level (positive event for tidymodels metrics)
    aki_stage2_7d = factor(aki_stage2_7d, levels = c(1, 0), 
                           labels = c("AKI", "No_AKI")),
    
    # Categorical predictors
    gender = factor(gender, levels = c("M", "F")),
    admission_type = factor(admission_type),
    race = factor(race, levels = race_levels),
    
    # Binary flags as integer
    across(c(dm, htn, ckd, copd, hf, cad, sepsis, vasopressor_24h), as.integer)
  ) |>
  select(stay_id, .imp, aki_stage2_7d, all_of(predictor_cols))

cat("Rows after dropping unknown outcomes:", nrow(stacked_clean), "\n")
cat("Unique stay_ids:", n_distinct(stacked_clean$stay_id), "\n")
```

### Outcome Distribution

```{r outcome-distribution}
stacked_clean |>
  filter(.imp == 1) |>
  count(aki_stage2_7d) |>
  mutate(pct = round(n / sum(n) * 100, 1)) |>
  kable(col.names = c("Outcome", "N", "%"),
        caption = "Outcome distribution (imputation 1, representative)")
```

### Residual Missingness Check

```{r missingness-check}
residual_na <- stacked_clean |>
  filter(.imp == 1) |>
  select(all_of(predictor_cols)) |>
  summarise(across(everything(), \(x) sum(is.na(x)))) |>
  pivot_longer(everything(), names_to = "variable", values_to = "n_missing") |>
  filter(n_missing > 0) |>
  arrange(desc(n_missing))

if (nrow(residual_na) == 0) {
  cat("No residual NAs in predictor columns. MICE imputation complete.\n")
} else {
  cat("WARNING — residual NAs found post-MICE:\n")
  print(residual_na)
}
```

---

## 4. Train/Test Split (80/20, Stratified on Unique Stay IDs)
The stacked data has 5 identical copies of each patient's outcome row (one per imputation). Splitting rows directly would allow the same patient's copies to appear in both train and test — data leakage. Splitting unique IDs first, then filtering the stacked data, ensures complete patient-level separation.

```{r train-test-split}
set.seed(203)

# Step 1: Get unique stay_ids with their outcomes
stay_outcomes <- stacked_clean |>
  filter(.imp == 1) |>
  select(stay_id, aki_stage2_7d)

# Step 2: 80/20 stratified split on unique stay_ids
split_ids <- initial_split(stay_outcomes, prop = 0.80, strata = aki_stage2_7d)
train_ids <- training(split_ids) |> pull(stay_id)
test_ids <- testing(split_ids)  |> pull(stay_id)

cat("Train stay_ids:", length(train_ids), "\n")
cat("Test stay_ids:", length(test_ids), "\n")

# Step 3: Filter stacked data
train_stack <- stacked_clean |> filter(stay_id %in% train_ids)
test_stack <- stacked_clean |> filter(stay_id %in% test_ids)

# Sanity checks
stopifnot(nrow(train_stack) == length(train_ids) * 5)
stopifnot(nrow(test_stack)  == length(test_ids)  * 5)
stopifnot(length(intersect(train_ids, test_ids)) == 0)

cat("Train rows:", nrow(train_stack), "(", length(train_ids), 
    "patients x 5 imputations)\n")
cat("Test rows:", nrow(test_stack), "(", length(test_ids), 
    "patients x 5 imputations)\n")
```

### Stratification Check

```{r stratification-check}
list(Train = train_ids, Test = test_ids) |>
  map_dfr(\(ids) {
    stay_outcomes |>
      filter(stay_id %in% ids) |>
      count(aki_stage2_7d) |>
      mutate(pct = round(n / sum(n) * 100, 1))
  }, .id = "Split") |>
  kable(col.names = c("Split", "Outcome", "N", "%"),
        caption = "Outcome distribution: train vs. test (unique stay_ids)")
```

---

## 5. Preprocessing Recipes

Two recipes: one for penalized regression and SVM (requires normalization), one for tree-based models.

The reference training frame uses `.imp == 1` to define column types and factor levels. The same recipe blueprint is applied to each imputed dataset at fit time.

```{r recipes}
# Using first imputation as reference for recipe prep (column types, factor levels)
train_imp1 <- train_stack |>
  filter(.imp == 1) |>
  select(-stay_id, -.imp)

# Recipe for Logistic Regression (LASSO) and SVM — normalize
recipe_glm <- recipe(aki_stage2_7d ~ ., data = train_imp1) |>
  step_dummy(gender, admission_type, race, one_hot = FALSE) |>
  step_zv(all_predictors()) |>
  step_normalize(all_numeric_predictors())

# Recipe for Random Forest and XGBoost — no normalization needed
recipe_tree <- recipe(aki_stage2_7d ~ ., data = train_imp1) |>
  step_dummy(gender, admission_type, race, one_hot = FALSE) |>
  step_zv(all_predictors())

# Verify recipes work on training data
cat("GLM recipe columns after prep:",
    ncol(prep(recipe_glm) |> bake(new_data = NULL)), "\n")
cat("Tree recipe columns after prep:",
    ncol(prep(recipe_tree) |> bake(new_data = NULL)), "\n")

# Check no NAs remain after recipe
baked_check <- prep(recipe_glm) |> bake(new_data = NULL)
if (any(is.na(baked_check))) {
  warning("NAs found after recipe baking — check predictor preparation")
} else {
  cat("No NAs in baked training data. \n")
}
```

---

## 6. Model Specifications

```{r model-specs}
n_cores <- max(1, parallel::detectCores() - 2)

# Model 0: Logistic Regression (unpenalized) — baseline
spec_logit <- logistic_reg() |>
  set_engine("glm") |>
  set_mode("classification")

# Model 1: Logistic Regression with LASSO (L1 penalty)
# mixture = 1 → pure LASSO; automatically selects features
spec_lr <- logistic_reg(
        penalty = tune(), 
        mixture = 1) |>
  set_engine("glmnet") |>
  set_mode("classification")

# Model 2: Random Forest 
spec_rf <- rand_forest(
        mtry = tune(), 
        min_n = tune(), 
        trees = 500) |>
  set_engine("ranger",
             importance = "impurity",
             num.threads = n_cores) |>
  set_mode("classification")

# Model 3: XGBoost
spec_xgb <- boost_tree(
  trees = tune(),
  tree_depth = tune(),
  learn_rate = tune(),
  min_n = tune(),
  loss_reduction = tune(),
  sample_size = tune()
) |>
  set_engine("xgboost") |>
  set_mode("classification")

# Model 4: Support Vector Machine (RBF kernel)
spec_svm <- svm_rbf(
        cost = tune(), 
        rbf_sigma = tune()) |>
  set_engine("kernlab") |>
  set_mode("classification")

# Logistic baseline workflow 
wf_logit <- workflow() |>
  add_recipe(recipe_glm) |>
  add_model(spec_logit)
```

---

## 7. Hyperparameter Tuning (on Imputation 1)

Tuning is performed on `.imp == 1` only. This is a deliberate performance-correctness trade-off: the variance in optimal hyperparameters across MICE imputations is typically small, and the computational cost of 5× tuning is not justified. The best hyperparameters from `.imp == 1` are then applied uniformly when fitting all 5 models.

```{r cv-setup}
# 5-fold CV stratified on outcome (within training set of imputation 1)
cv_folds <- vfold_cv(train_imp1, v = 5, strata = aki_stage2_7d)
tune_metric <- metric_set(roc_auc)
```

### 7.1 Logistic Regression (LASSO)

```{r tune-lr}
wf_lr <- workflow() |>
  add_recipe(recipe_glm) |>
  add_model(spec_lr)

lr_grid <- grid_regular(
  penalty(range = c(-4, 0), trans = log10_trans()),
  levels = 30
)

cl <- makePSOCKcluster(n_cores)
registerDoParallel(cl)

tune_lr <- tune_grid(
  wf_lr,
  resamples = cv_folds,
  grid = lr_grid,
  metrics = tune_metric,
  control = control_grid(save_pred = FALSE, verbose = FALSE)
)

stopCluster(cl)

best_lr <- select_best(tune_lr, metric = "roc_auc")
cat("Best LASSO penalty:", best_lr$penalty, "\n")
cat("Best CV AUROC (LASSO):", round(show_best(tune_lr, metric = "roc_auc", 
                                              n = 1)$mean, 4), "\n")

autoplot(tune_lr) +
  labs(title = "LASSO: CV AUROC vs. Penalty") +
  theme_minimal()
```

### 7.2 Random Forest

```{r tune-rf}
wf_rf <- workflow() |>
  add_recipe(recipe_tree) |>
  add_model(spec_rf)

rf_grid <- grid_latin_hypercube(
  mtry(range = c(5L, 25L)),
  min_n(range = c(5L, 30L)),
  size = 20
)

cl <- makePSOCKcluster(n_cores)
registerDoParallel(cl)

tune_rf <- tune_grid(
  wf_rf,
  resamples = cv_folds,
  grid = rf_grid,
  metrics = tune_metric,
  control = control_grid(save_pred = FALSE, verbose = FALSE)
)

stopCluster(cl)

best_rf <- select_best(tune_rf, metric = "roc_auc")
cat("Best RF mtry:", best_rf$mtry, "| min_n:", best_rf$min_n, "\n")
cat("Best CV AUROC (RF):", round(show_best(tune_rf, metric = "roc_auc", 
                                           n = 1)$mean, 4), "\n")

autoplot(tune_rf) +
  labs(title = "Random Forest: CV AUROC vs. Hyperparameters") +
  theme_minimal()
```

### 7.3 XGBoost

```{r tune-xgb}
wf_xgb <- workflow() |>
  add_recipe(recipe_tree) |>
  add_model(spec_xgb)

xgb_grid <- grid_latin_hypercube(
  trees(range = c(100L, 1000L)),
  tree_depth(range = c(3L, 8L)),
  learn_rate(range = c(-3, -1)),
  min_n(range = c(5L, 30L)),
  loss_reduction(),
  sample_size = sample_prop(range = c(0.5, 1.0)),
  size = 30
)

cl <- makePSOCKcluster(n_cores)
registerDoParallel(cl)

tune_xgb <- tune_grid(
  wf_xgb,
  resamples = cv_folds,
  grid = xgb_grid,
  metrics = tune_metric,
  control = control_grid(save_pred = FALSE, verbose = FALSE)
)

stopCluster(cl)

best_xgb <- select_best(tune_xgb, metric = "roc_auc")
cat("Best XGB trees:", best_xgb$trees,
    "| depth:", best_xgb$tree_depth,
    "| lr:", round(best_xgb$learn_rate, 5), "\n")
cat("Best CV AUROC (XGBoost):", round(show_best(tune_xgb, metric = "roc_auc", 
                                                n = 1)$mean, 4), "\n")

autoplot(tune_xgb) +
  labs(title = "XGBoost: CV AUROC vs. Hyperparameters") +
  theme_minimal()
```

### 7.4 SVM (RBF Kernel)

```{r tune-svm}
wf_svm <- workflow() |>
  add_recipe(recipe_glm) |>
  add_model(spec_svm)

svm_grid <- grid_latin_hypercube(
  cost(),
  rbf_sigma(),
  size = 20
)

cl <- makePSOCKcluster(n_cores)
registerDoParallel(cl)

tune_svm <- tune_grid(
  wf_svm,
  resamples = cv_folds,
  grid = svm_grid,
  metrics = tune_metric,
  control = control_grid(save_pred = FALSE, verbose = FALSE)
)

stopCluster(cl)

best_svm <- select_best(tune_svm, metric = "roc_auc")
cat("Best SVM cost:", round(best_svm$cost, 4),
    "| rbf_sigma:", round(best_svm$rbf_sigma, 6), "\n")
cat("Best CV AUROC (SVM):", round(show_best(tune_svm, metric = "roc_auc", 
                                            n = 1)$mean, 4), "\n")

autoplot(tune_svm) +
  labs(title = "SVM: CV AUROC vs. Hyperparameters") +
  theme_minimal()
```

### CV Results Summary

```{r cv-summary}
# Logistic baseline: fit inline on train_imp1 
logit_train_auroc <- fit(wf_logit, data = train_imp1) |>
  augment(train_imp1) |>
  roc_auc(truth = aki_stage2_7d, .pred_AKI, event_level = "first") |>
  pull(.estimate)

logit_cv_row <- tibble(
  model = "Logistic Regression",
  mean = logit_train_auroc,
  std_err = NA_real_,
  n = NA_integer_
)

cv_summary <- bind_rows(
  logit_cv_row,
  show_best(tune_lr,  metric = "roc_auc", n = 1) |> mutate(model = "LASSO"),
  show_best(tune_rf,  metric = "roc_auc", n = 1) |> mutate(model = "Random Forest"),
  show_best(tune_xgb, metric = "roc_auc", n = 1) |> mutate(model = "XGBoost"),
  show_best(tune_svm, metric = "roc_auc", n = 1) |> mutate(model = "SVM")
) |>
  select(model, mean, std_err, n) |>
  arrange(desc(mean))

cv_summary |>
  kable(digits = 4,
        col.names = c("Model", "CV/Train AUROC", "Std. Error", "N folds"),
        caption = "Cross-validation performance on imputation 1 training set 
        (Logistic Regression = training AUROC, no CV)")
```

---

## 8. Fit 5 Models per Algorithm (MI-Correct)

Using best hyperparameters from Section 7, fit each model on all 5 imputed training sets.

```{r finalize-workflows}
# wf_logit already defined in model-specs chunk above
final_wf_lr  <- finalize_workflow(wf_lr,  best_lr)
final_wf_rf  <- finalize_workflow(wf_rf,  best_rf)
final_wf_xgb <- finalize_workflow(wf_xgb, best_xgb)
final_wf_svm <- finalize_workflow(wf_svm, best_svm)
```

```{r fit-all-imputations}
# Fit one workflow on one imputed training dataset
fit_on_imp <- function(wf, imp_num) {
  train_i <- train_stack |>
    filter(.imp == imp_num) |>
    select(-stay_id, -.imp)
  fit(wf, data = train_i)
}

# Fit 5 x Logistic Regression
fits_logit <- map(1:5, \(i) fit_on_imp(wf_logit,i))

# Fit 5 x LASSO models
fits_lr <- map(1:5, \(i) fit_on_imp(final_wf_lr,i))

# Fit 5 x Random Forest models
fits_rf <- map(1:5, \(i) fit_on_imp(final_wf_rf,i))

# Fitt 5 x XGBoost models
fits_xgb <- map(1:5, \(i) fit_on_imp(final_wf_xgb,i))

# Fit 5 x SVM models
fits_svm <- map(1:5, \(i) fit_on_imp(final_wf_svm,  i))

```

---

## 9. Pooled Predictions on Test Set

For each patient in the test set, we generate 5 predicted probabilities (one per imputed test dataset) and average them. This implements Rubin's rule adapted for predicted probabilities.

```{r pool-predictions}
# True outcomes from test set (identical across all .imp)
test_true <- test_stack |>
  filter(.imp == 1) |>
  select(stay_id, aki_stage2_7d)

# Get predicted P(AKI) from imputation i using a list of 5 fitted models
get_test_probs <- function(fit_list, imp_num) {
  test_i <- test_stack |>
    filter(.imp == imp_num) |>
    select(-stay_id, -.imp)

  predict(fit_list[[imp_num]], test_i, type = "prob") |>
    pull(.pred_AKI)
}

# Average predicted probs across 5 imputations
pool_predictions <- function(fit_list, model_name) {
  prob_mat <- map(1:5, \(i) get_test_probs(fit_list, i)) |>
    bind_cols() |>
    setNames(paste0("imp", 1:5))

  test_true |>
    bind_cols(prob_mat) |>
    mutate(
      .pred_AKI = rowMeans(across(starts_with("imp"))),
      model = model_name
    ) |>
    select(stay_id, aki_stage2_7d, .pred_AKI, model)
}

pred_logit <- pool_predictions(fits_logit, "Logistic Regression")
pred_lr <- pool_predictions(fits_lr,    "LASSO")
pred_rf <- pool_predictions(fits_rf,    "Random Forest")
pred_xgb <- pool_predictions(fits_xgb,  "XGBoost")
pred_svm <- pool_predictions(fits_svm,  "SVM")

all_preds <- bind_rows(pred_logit, pred_lr, pred_rf, pred_xgb, pred_svm)

cat("Pooled predictions shape:", nrow(all_preds), "rows x", 
    ncol(all_preds), "cols\n")
cat("Expected:", n_distinct(test_true$stay_id) * 5, 
    "rows (test patients × 5 models)\n")
```

---

## 9b. DeLong's Test: XGBoost vs Logistic Regression

```{r delong_test}
# Extract pooled predictions per model
preds_xgb <- all_preds |> filter(model == "XGBoost")
preds_logit <- all_preds |> filter(model == "Logistic Regression")

# Build ROC objects (response: 1 = AKI, 0 = No AKI)
true_labels_xgb <- as.numeric(preds_xgb$aki_stage2_7d == "AKI")
true_labels_logit <- as.numeric(preds_logit$aki_stage2_7d == "AKI")

roc_xgb <- pROC::roc(true_labels_xgb,   preds_xgb$.pred_AKI, quiet = TRUE)
roc_logit <- pROC::roc(true_labels_logit, preds_logit$.pred_AKI, quiet = TRUE)

# DeLong's test (two-sided)
delong_result <- pROC::roc.test(roc_xgb, roc_logit, method = "delong")

cat("XGBoost AUROC: ", round(auc(roc_xgb), 4), "\n")
cat("Logistic Regression AUROC: ", round(auc(roc_logit), 4), "\n")
cat("DeLong's test Z =", round(delong_result$statistic, 3),
    "  P =", round(delong_result$p.value, 4), "\n")
```

---

## 10. Evaluation Metrics

### 10.1 AUROC

```{r auroc}
roc_results <- all_preds |>
  group_by(model) |>
  roc_auc(truth = aki_stage2_7d, .pred_AKI, event_level = "first") |>
  rename(AUROC = .estimate) |>
  select(model, AUROC) |>
  arrange(desc(AUROC))

roc_results |>
  kable(digits = 4, caption = "AUROC by model 
        (test set, pooled MI predictions)")
```

### 10.2 AUPRC (Precision-Recall AUC)

```{r auprc}
pr_results <- all_preds |>
  group_by(model) |>
  pr_auc(truth = aki_stage2_7d, .pred_AKI, event_level = "first") |>
  rename(AUPRC = .estimate) |>
  select(model, AUPRC)

pr_results |>
  kable(digits = 4, caption = "AUPRC by model")
```

### 10.3 Brier Score

```{r brier}
brier_results <- all_preds |>
  mutate(outcome_bin = as.integer(aki_stage2_7d == "AKI")) |>
  group_by(model) |>
  summarise(
    Brier = round(mean((.pred_AKI - outcome_bin)^2), 4),
    .groups = "drop"
  )

brier_results |>
  kable(digits = 4, caption = "Brier Score by model 
        (lower = better; 0.25 = uninformative)")
```

### 10.4 Optimal Threshold via Youden's J

```{r threshold-youden}
threshold_results <- all_preds |>
  group_by(model) |>
  threshold_perf(
    truth = aki_stage2_7d,
    estimate = .pred_AKI,
    thresholds = seq(0.05, 0.95, by = 0.01),
    event_level = "first",
    metrics = metric_set(sens, spec, j_index)
  ) |>
  filter(.metric == "j_index") |>
  group_by(model) |>
  slice_max(.estimate, n = 1, with_ties = FALSE) |>
  select(model, optimal_threshold = .threshold, j_index = .estimate)

threshold_results |>
  kable(digits = 3, caption = "Optimal classification threshold by Youden's J")
```

### 10.5 Confusion Matrix at Optimal Threshold

```{r confusion-matrix}
preds_with_threshold <- all_preds |>
  left_join(threshold_results, by = "model") |>
  mutate(
    .pred_class = factor(
      if_else(.pred_AKI >= optimal_threshold, "AKI", "No_AKI"),
      levels = c("AKI", "No_AKI")
    )
  )

# Print confusion matrix for each model
for (mod in unique(all_preds$model)) {
  cat("\n--- Confusion Matrix:", mod, "---\n")
  cm <- preds_with_threshold |>
    filter(model == mod) |>
    conf_mat(truth = aki_stage2_7d, estimate = .pred_class)
  print(cm)
  cat("Sensitivity:", round(cm$table[1,1] / (cm$table[1,1] + 
                                                     cm$table[2,1]), 3), "\n")
  cat("Specificity:", round(cm$table[2,2] / (cm$table[1,2] + 
                                                     cm$table[2,2]), 3), "\n")
}
```

### 10.6 Combined Performance Table

```{r performance-table}
performance_table <- roc_results |>
  left_join(pr_results, by = "model") |>
  left_join(brier_results, by = "model") |>
  left_join(threshold_results, by = "model") |>
  arrange(desc(AUROC))

performance_table |>
  kable(
    digits = 4,
    col.names = c("Model", "AUROC", "AUPRC", "Brier Score",
                  "Optimal Threshold", "Youden's J"),
    caption = "Table 1: ML Model Performance on Held-Out Test Set 
    (Pooled MI Predictions)"
  )
```

---

## 11. Figures

### Figure 1: ROC Curves

```{r fig-roc, fig.height=6, fig.width=8}
roc_curves <- all_preds |>
  group_by(model) |>
  roc_curve(truth = aki_stage2_7d, .pred_AKI, event_level = "first")

roc_label <- roc_results |>
  mutate(label = paste0(model, " (AUC = ", round(AUROC, 3), ")")) |>
  select(model, label)

roc_plot <- roc_curves |>
  left_join(roc_label, by = "model") |>
  ggplot(aes(x = 1 - specificity, y = sensitivity, color = label)) +
  geom_path(linewidth = 1.0) +
  geom_abline(linetype = "dashed", color = "grey60") +
  scale_color_brewer(palette = "Set1", name = NULL) +
  coord_equal() +
  labs(
    title = "ROC Curves — AKI Stage 2+ Prediction",
    subtitle = "Predictions pooled across 5 MICE imputations",
    x = "1 - Specificity (False Positive Rate)",
    y = "Sensitivity (True Positive Rate)"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = c(0.65, 0.2),
    legend.background = element_rect(fill = "white", color = NA),
    plot.title = element_text(face = "bold")
  )

roc_plot
```

### Figure 2: Precision-Recall Curves

```{r fig-pr, fig.height=6, fig.width=8}
pr_curves <- all_preds |>
  group_by(model) |>
  pr_curve(truth = aki_stage2_7d, .pred_AKI, event_level = "first")

pr_label <- pr_results |>
  mutate(label = paste0(model, " (AUPRC = ", round(AUPRC, 3), ")")) |>
  select(model, label)

# Baseline: outcome prevalence
prev <- mean(all_preds$aki_stage2_7d[all_preds$model == "LASSO"] == "AKI")

pr_plot <- pr_curves |>
  left_join(pr_label, by = "model") |>
  ggplot(aes(x = recall, y = precision, color = label)) +
  geom_path(linewidth = 1.0) +
  geom_hline(yintercept = prev, linetype = "dashed", color = "grey60") +
  annotate("text", x = 0.85, y = prev + 0.02,
           label = paste0("Baseline (", round(prev * 100, 1), "%)"),
           size = 3.5, color = "grey40") +
  scale_color_brewer(palette = "Set1", name = NULL) +
  coord_cartesian(xlim = c(0, 1), ylim = c(0, 1)) +
  labs(
    title = "Precision-Recall Curves — AKI Stage 2+ Prediction",
    subtitle = "Dashed line = no-skill baseline (outcome prevalence)",
    x = "Recall (Sensitivity)",
    y = "Precision (PPV)"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = c(0.7, 0.8),
    legend.background = element_rect(fill = "white", color = NA),
    plot.title = element_text(face = "bold")
  )

pr_plot
```

### Figure 3: Calibration Plots

```{r fig-calibration, fig.height=8, fig.width=10}
calibration_df <- all_preds |>
  mutate(outcome_bin = as.integer(aki_stage2_7d == "AKI")) |>
  group_by(model) |>
  mutate(bin = cut(.pred_AKI,
                   breaks = seq(0, 1, by = 0.1),
                   include.lowest = TRUE, right = FALSE)) |>
  group_by(model, bin) |>
  summarise(
    mean_pred = mean(.pred_AKI),
    mean_obs = mean(outcome_bin),
    n = n(),
    .groups = "drop"
  ) |>
  filter(n >= 10)

ggplot(calibration_df, aes(x = mean_pred, y = mean_obs)) +
  geom_abline(linetype = "dashed", color = "grey60") +
  geom_point(aes(size = n), color = "#2c7bb6", alpha = 0.8) +
  geom_line(color = "#2c7bb6", alpha = 0.6) +
  scale_size_continuous(range = c(2, 8), name = "N in bin") +
  coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +
  facet_wrap(~ model, ncol = 2) +
  labs(
    title = "Calibration Plots — Predicted vs. Observed Probability",
    subtitle = "Each point = decile bin; dashed = perfect calibration",
    x = "Mean Predicted Probability",
    y = "Observed Event Rate"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

### Figure 4: LASSO — Top Non-Zero Coefficients

```{r fig-lasso-coefs, fig.height=7, fig.width=9}
lasso_coefs <- fits_lr[[1]] |>
  extract_fit_parsnip() |>
  tidy() |>
  filter(term != "(Intercept)", estimate != 0) |>
  mutate(direction = if_else(estimate > 0, "Positive (risk)", 
                             "Negative (protective)")) |>
  arrange(desc(abs(estimate))) |>
  slice_head(n = 25)

ggplot(lasso_coefs, aes(x = reorder(term, abs(estimate)), y = estimate, 
                        fill = direction)) +
  geom_col(show.legend = TRUE) +
  scale_fill_manual(values = c("Positive (risk)" = "#e74c3c",
                               "Negative (protective)" = "#3498db"),
                    name = NULL) +
  coord_flip() +
  labs(
    title = "LASSO: Top Non-Zero Coefficients",
    subtitle = "Imputation 1; positive = increased AKI risk",
    x = NULL,
    y = "Coefficient Estimate"
  ) +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"),
        legend.position = "bottom")
```

### Figure 5: Random Forest — Feature Importance

```{r fig-rf-importance, fig.height=7, fig.width=9}
rf_importance <- map_dfr(1:5, function(i) {
  fits_rf[[i]] |>
    extract_fit_parsnip() |>
    pluck("fit") |>
    (\(m) tibble(
      variable = names(m$variable.importance),
      importance = m$variable.importance
    ))() |>
    mutate(.imp = i)
}) |>
  group_by(variable) |>
  summarise(mean_importance = mean(importance), .groups = "drop") |>
  arrange(desc(mean_importance)) |>
  slice_head(n = 25)

ggplot(rf_importance, aes(x = reorder(variable, mean_importance), 
                          y = mean_importance)) +
  geom_col(fill = "#27ae60") +
  coord_flip() +
  labs(
    title = "Random Forest: Top 25 Features by Mean Gini Importance",
    subtitle = "Averaged across 5 MICE imputations",
    x = NULL,
    y = "Mean Impurity (Gini) Importance"
  ) +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"))
```

### Figure 6: XGBoost — Feature Importance (Gain)

```{r fig-xgb-importance, fig.height=7, fig.width=9}
xgb_importance <- map_dfr(1:5, function(i) {
  fits_xgb[[i]] |>
    extract_fit_parsnip() |>
    pluck("fit") |>
    xgb.importance(model = _) |>
    as_tibble() |>
    mutate(.imp = i)
}) |>
  group_by(Feature) |>
  summarise(mean_gain = mean(Gain), .groups = "drop") |>
  arrange(desc(mean_gain)) |>
  slice_head(n = 25)

ggplot(xgb_importance, aes(x = reorder(Feature, mean_gain), y = mean_gain)) +
  geom_col(fill = "#e67e22") +
  coord_flip() +
  labs(
    title = "XGBoost: Top 25 Features by Mean Gain",
    subtitle = "Averaged across 5 MICE imputations",
    x = NULL,
    y = "Mean Gain"
  ) +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"))
```

---

## 12. Save Outputs

```{r save-outputs}
saveRDS(all_preds, "ml_predictions_pooled.rds")
write_csv(performance_table, "ml_performance_table.csv")
saveRDS(
  list(logit = fits_logit, lr = fits_lr, rf = fits_rf, xgb = fits_xgb, 
       svm = fits_svm),
  "ml_fitted_models.rds"
)

cat("Saved:\n")
cat("  ml_predictions_pooled.rds\n")
cat("  ml_performance_table.csv\n")
cat("  ml_fitted_models.rds\n")
```

---

## Session Info

```{r session-info}
sessionInfo()
```
